---
title: "DCRW examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DCRW examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
csl: apa-old-doi-prefix.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

> This page shows how the examples in Sections 5.1 and 5.2 of @shankar2025 are generated.

```{r setup, message=FALSE}
library(roams)
library(janitor)
library(tsibble)
library(tidyverse)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(knitr)

# Get country boundaries as an sf object for later use
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
```

These examples all use the first-differenced correlated random walk (DCRW) model, a type of linear Gaussian SSM. This model is detailed in @auger-methe_guide_2021. The DCRW model has the following observation and state processes,
\begin{align} \label{eq:DCRW_equations}
\begin{split}
    \mathbf{y}_t &= \mathbf{z}_{t} + \mathbf{v}_t \\
    \mathbf{z}_t &= \mathbf{z}_{t-1} + \phi(\mathbf{z}_{t-1} - \mathbf{z}_{t-2}) + \mathbf{w}_t,
\end{split}
\end{align}
where $\mathbf{y}_t$ is the location observed by the satellite, $\mathbf{v}_t$ is the error from the satellite measurements, $\mathbf{z}_t$ is the true location of the whale, and $\mathbf{w}_t$ captures the randomness in the whale's movement. Both $\mathbf{y}_t$ and $\mathbf{z}_t$ are 2-dimensional vectors containing longitude and latitude coordinates. The DCRW model assumes that the whale's location depends on the previous location $\mathbf{z}_{t-1}$, and the previous direction and speed of movement $\mathbf{z}_{t-1} - \mathbf{z}_{t-2}$. The amount of dependence on $\mathbf{z}_{t-1} - \mathbf{z}_{t-2}$ is determined by the correlation parameter $\phi\in [0,1)$.

Notice the use of $\mathbf{z}$ instead of $\mathbf{x}$ for the state process; the DCRW model above is not in state-space form since the state process depends on the previous \textit{two} states, $\mathbf{z}_{t-1}$ and $\mathbf{z}_{t-2}$. This presentation makes it easier to understand the model dynamics, but in order to estimate the parameters of the model, the model must be converted into state-space form. See the beginning of Section 4 of @shankar2025 for a summary on how this is done.

## Animals {.tabset}

We illustrate our method on five example data sets.

### Cats

```{r}
cat_mia = read_csv("data/Feral cat (Felis catus) - Scotia, NSW.csv") |> 
  clean_names() |> 
  mutate(algorithm_marked_outlier = ifelse(is.na(algorithm_marked_outlier), 
                                                 FALSE,
                                                 algorithm_marked_outlier),
         manually_marked_outlier = ifelse(is.na(manually_marked_outlier), 
                                                 FALSE,
                                                 manually_marked_outlier),
         outlier = algorithm_marked_outlier | manually_marked_outlier) |> 
  filter(individual_local_identifier == "Mia_FC642") |> 
  mutate(timestamp = lubridate::round_date(timestamp, "20 minutes")) |> 
  tsibble::tsibble() |>
  tsibble::fill_gaps() |> 
  slice(-(1:19)) |> 
  as_tibble() |> 
  rename(x = location_long,
         y = location_lat)

cat_max = read_csv("data/Feral cat (Felis catus) - Scotia, NSW.csv") |> 
  clean_names() |> 
  mutate(algorithm_marked_outlier = ifelse(is.na(algorithm_marked_outlier), 
                                                 FALSE,
                                                 algorithm_marked_outlier),
         manually_marked_outlier = ifelse(is.na(manually_marked_outlier), 
                                                 FALSE,
                                                 manually_marked_outlier),
         outlier = algorithm_marked_outlier | manually_marked_outlier) |> 
  filter(individual_local_identifier == "Max_MC629") |> 
  mutate(timestamp = lubridate::round_date(timestamp, "20 minutes")) |> 
  tsibble::tsibble() |>
  tsibble::fill_gaps() |> 
  slice(-(1:24)) |> 
  as_tibble() |> 
  rename(x = location_long,
         y = location_lat)
```

Plot with world map data:

```{r raw-data-vis-cat, fig.height=5}
# Work out x/y ranges with some padding
x_range <- range(cat_mia$x, na.rm = TRUE)
y_range <- range(cat_mia$y, na.rm = TRUE)

x_pad <- diff(x_range) * 0.05  # 5% padding
y_pad <- diff(y_range) * 0.05 # 5% padding

p1 = cat_mia |> 
  ggplot() +
  # Country boundaries
  geom_sf(data = world, fill = "grey95", colour = "grey70", size = 0.3) +
  # Animal track and points
  geom_path(aes(x = x, y = y, colour = timestamp), alpha = 0.2) +
  geom_point(aes(x = x, y = y, colour = timestamp)) +
  # Colour scale
  viridis::scale_color_viridis() +
  # Labels and theme
  labs(x = "Longitude", y = "Latitude", title = "Cat") +
  theme_minimal() +
  theme(legend.position = "none") +
  # Focus map on whale’s range
  coord_sf(
    xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
    ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
    expand = FALSE
  )
p1
```

### Cows

```{r}
cow = read_csv("data/export_UQ_GPS.csv") |> 
  clean_names() |> 
  filter(partition_key == "70741400000EB036") |> 
  # The cow data longitude and latitude are swapped! So we set x to be latitude and y to be longitude:
  rename(x = latitude,
         y = longitude) |> 
  mutate(date = date(timestamp)) |> 
  select(date, timestamp, x, y) |>
  group_by(date) |> 
  mutate(diff = c(2.38, diff(timestamp))) |>
  mutate(missing_rows = map(diff, ~ seq_len(round(. / 2.38)))) |>
  unnest(missing_rows) |>
  mutate(x = ifelse(missing_rows > 1, NA, x),
         y = ifelse(missing_rows > 1, NA, y)) |> 
  ungroup() |> 
  rename(x = x,
         y = y)

cow_11 = cow |> filter(date == "2024-11-11")
```

Plot with world map data:

```{r raw-data-vis-cow, fig.height=5}
# Work out x/y ranges with some padding
x_range <- range(cow_11$x, na.rm = TRUE)
y_range <- range(cow_11$y, na.rm = TRUE)

x_pad <- diff(x_range) * 0.05  # 5% padding
y_pad <- diff(y_range) * 0.05 # 5% padding

p1 %+% cow_11 + 
  # Focus map on s's range
  coord_sf(
    xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
    ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
    expand = FALSE
  ) + 
  labs(title = "Cow 11")
```

### Polar bear

```{r polar-bear-import}
polar_low_freq = read_csv("data/PB_Argos.csv") |> 
  # Standardise column names
  clean_names() |> 
  # Convert date_time column to a proper datetime object called 'timestamp'
  mutate(timestamp = ymd_hms(date_time)) |> 
  # Drop the original date_time column
  select(-date_time) |> 
  # Round timestamps to the nearest day
  mutate(timestamp = lubridate::round_date(timestamp, "1 day")) |> 
  # Within each group, keep only the first row (e.g. earliest observation per day)
  group_by(timestamp) |> 
  slice_head(n = 1) |> 
  # Convert timestamp to a Date object (drop time-of-day information)
  mutate(timestamp = as.Date(timestamp)) |> 
  # Convert the data to a tsibble object (time series tibble)
  tsibble::tsibble() |>
  # Fill in missing time points in the series with explicit gaps
  tsibble::fill_gaps() |>
  # Convert back to a regular tibble
  as_tibble() |> 
  # Rename longitude/latitude columns to x and y
  rename(x = lon, y = lat)
```


Plot with world map data:

```{r raw-data-vis-polar_bear}
# Work out x/y ranges with some padding
x_range <- range(polar_low_freq$x, na.rm = TRUE)
y_range <- range(polar_low_freq$y, na.rm = TRUE)

x_pad <- diff(x_range) * 0.05  # 5% padding
y_pad <- diff(y_range) * 0.05 # 5% padding

p2 = p1 %+% polar_low_freq + 
  # Focus map on polar bear's range
  coord_sf(
    xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
    ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
    expand = FALSE
  ) + 
  labs(title = "Polar bear")
p2
```



### Seal

```{r seal-import}
seal = read_csv("data/sealLocs.csv") |> 
  # Keep only rows for the seal with ID "stephanie"
  filter(id == "stephanie") |> 
  # Round timestamps to the nearest day
  mutate(timestamp = lubridate::round_date(time, "1 day")) |> 
  # Within each group, keep only the last row (e.g. latest observation per day)
  group_by(timestamp) |> 
  slice_tail(n = 1) |> 
  # Convert timestamp to a Date object (drop time-of-day information)
  mutate(timestamp = as.Date(timestamp)) |> 
  # Convert the data to a tsibble object indexed by timestamp
  tsibble::tsibble(index = timestamp) |>
  # Fill in missing time points in the series with explicit gaps
  tsibble::fill_gaps() |>
  # Convert back to a regular tibble
  as_tibble() |> 
  # Rename longitude/latitude columns to x and y
  rename(x = lon,
         y = lat)
```


Plot with world map data:

```{r raw-data-vis-seal, fig.height=5}
# Work out x/y ranges with some padding
x_range <- range(seal$x, na.rm = TRUE)
y_range <- range(seal$y, na.rm = TRUE)

x_pad <- diff(x_range) * 0.05  # 5% padding
y_pad <- diff(y_range) * 0.05 # 5% padding

p1 %+% seal + 
  # Focus map on s's range
  coord_sf(
    xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
    ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
    expand = FALSE
  ) + 
  labs(title = "Seal")
```

### Blue whale

```{r whale-2008-import}
whale_2008 = read_csv("data/Blue whales Eastern North Pacific 1993-2008 - Argos Data.csv") |>
  # Standardise column names
  janitor::clean_names() |>  
  # Keep only rows for the whale with ID "2008CA-Bmu-10839"
  filter(individual_local_identifier == "2008CA-Bmu-10839") |> 
  # Replace missing values in 'manually_marked_outlier' with FALSE
  mutate(manually_marked_outlier = tidyr::replace_na(manually_marked_outlier, FALSE)) |> 
  # Round timestamps to the nearest 12 hours
  mutate(timestamp = lubridate::round_date(timestamp, "12 hours")) |> 
  # Within each group, keep only the last row (e.g. latest observation per 12-hour block)
  group_by(timestamp) |> 
  slice_tail(n = 1) |> 
  # Convert the data to a tsibble object (time series tibble)
  tsibble::tsibble() |>
  # Fill in missing time points in the series with explicit gaps
  tsibble::fill_gaps() |>
   # Convert back to a regular tibble
  as_tibble() |> 
  # Rename longitude/latitude columns to x and y
  rename(x = location_long,
         y = location_lat)
```

Plot with world map data:

```{r raw-data-vis-whale}
# Work out x/y ranges with some padding
x_range <- range(whale_2008$x, na.rm = TRUE)
y_range <- range(whale_2008$y, na.rm = TRUE)

x_pad <- diff(x_range) * 0.05  # 5% padding
y_pad <- diff(y_range) * 0.05 # 5% padding

p1 %+% whale_2008 + 
  # Focus map on s's range
  coord_sf(
    xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
    ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
    expand = FALSE
  ) + 
  labs(title = "Blue whale")
```

## Data summary

Combine the data sets into a list:

```{r}
data_sets = list(
  "cat_mia" = cat_mia,
  "cow_11" = cow_11,
  "polar_low_freq" = polar_low_freq,
  "seal" = seal,
  "whale_2008" = whale_2008
)
```

### Summary of missingness proportion and size of data sets

```{r}
# Add column containing frequency of measurements for each data set
frequencies = c("20 minutes", 
                "2 minutes and 23 seconds", 
                "1 day", "1 day", "12 hours")
data_sets |> 
  map_dbl(~ mean(is.na(.$x))) |> 
  round(2) |> 
  as_tibble(rownames = "data_set") |> 
  rename(proportion_missing = value) |> 
  mutate(n = data_sets |> map_dbl(~ length(.$x))) |> 
  arrange(data_set) |> 
  select(data_set, n, proportion_missing) |> 
  mutate(interval = frequencies) |>
  knitr::kable(
    col.names = c("Data set", "Number of observations",  
                  "Proportion missing", "Interval"))
```

## Model fitting

The last 10% of points from each data set are left out to use as a 'test set'. The other 90% are used as a 'training set'.

The code below fits the classical, ROAMS, Huber and trimmed methods to the training sets. It also computes the filtered and predicted values on the test sets so that mean-squared forecast errors (MSFE) can be reported.

```{r eval = FALSE}
results = tibble(
  name = character(),
  data_set = list(),
  y_train = list(),
  y_oos = list(),
  
  classical_model = list(),
  roams_model = list(),
  roams_model_list = list(),
  huber_model = list(),
  trimmed_model = list(),
  
  classical_oos = list(),
  roams_fut_oos = list(),
  roams_kalman_oos = list(),
  huber_oos = list(),
  trimmed_oos = list(),
  
  oos_outliers_flagged = list()
)

for (i in 1:length(data_sets)) {
  
  y_all = data_sets[[i]] |> 
    select(x, y) |> 
    as.matrix()
  
  n = nrow(y_all)
  n_train = round(0.9*n)  # 90%/10% train/test split
  n_oos = n - n_train
  
  y_train = y_all[1:n_train,]
  y_oos = y_all[(n_train+1):n,]
  
  var_est = c(mad(diff(y_train[,1]), na.rm = TRUE)^2,
              mad(diff(y_train[,2]), na.rm = TRUE)^2)
  
  build_fn = function(par) {
    
    phi_coef = par[1]
    Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
    Phi[1,3] = -phi_coef
    Phi[2,4] = -phi_coef
    Phi[3,1] = 1
    Phi[4,2] = 1
    
    A = diag(4)[1:2,]
    Q = diag(c(par[2], par[3], 0, 0))
    R = diag(c(par[4], par[5]))
    
    x0 = c(y_train[1,], y_train[1,])
    P0 = diag(rep(0, 4))
    
    specify_SSM(
      state_transition_matrix = Phi,
      state_noise_var = Q,
      observation_matrix = A,
      observation_noise_var = R,
      init_state_mean = x0,
      init_state_var = P0)
  }
  
  build_fn_huber_trimmed = function(par) {
    
    phi_coef = par[1]
    Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
    Phi[1,3] = -phi_coef
    Phi[2,4] = -phi_coef
    Phi[3,1] = 1
    Phi[4,2] = 1
    
    A = diag(4)[1:2,]
    Q = diag(c(par[2], par[3], 0, 0))
    R = diag(c(par[4], par[5]))
    
    second_complete_obs = which(!is.na(y_train[,1]))[2]
    diff = sum(abs(y_train[1,] - y_train[second_complete_obs,]))
    x0 = c(y_train[1,], y_train[1,] + diff * 0.1)
    P0 = diag(rep(0, 4))
    
    specify_SSM(
      state_transition_matrix = Phi,
      state_noise_var = Q,
      observation_matrix = A,
      observation_noise_var = R,
      init_state_mean = x0,
      init_state_var = P0)
  }
  
  classical_model = classical_SSM(
    y = y_train, 
    init_par = c(0.5, var_est, var_est), 
    build = build_fn, 
    lower = c(0, rep(1e-12, 4)),
    upper = c(1, rep(Inf, 4)),
  )
  
  roams_model_list = roams_SSM(
    y = y_train,
    init_par = c(0.5, var_est, var_est),
    build = build_fn,
    num_lambdas = 50,
    cores = 4,
    lower = c(0, rep(1e-12, 4)),
    upper = c(1, rep(Inf, 4)),
    B = 50
  )
  
  huber_model = huber_robust_SSM(
    y = y_train,
    init_par = c(0.5, var_est, var_est),
    build = build_fn_huber_trimmed,
    lower = c(0.001, rep(1e-12, 4)),
    upper = c(0.999, rep(Inf, 4))
  )
  
  trimmed_model = trimmed_robust_SSM(
    y = y_train,
    init_par = c(0.5, var_est, var_est),
    build = build_fn_huber_trimmed,
    alpha = 0.05,  # assume a low level of 5% contamination
    lower = c(0.001, rep(1e-12, 4)),
    upper = c(0.999, rep(Inf, 4))
  )
  
  classical_model = attach_insample_info(classical_model)
  huber_model = attach_insample_info(huber_model)
  trimmed_model = attach_insample_info(trimmed_model)
  
  roams_model = best_BIC_model(roams_model_list)
  roams_model = attach_insample_info(roams_model)
  
  # Out-of-sample metrics
  
  # Re-create build function with different initial state mean
  build_oos_fn = function(par) {

    phi_coef = par[1]
    Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
    Phi[1,3] = -phi_coef
    Phi[2,4] = -phi_coef
    Phi[3,1] = 1
    Phi[4,2] = 1

    A = diag(4)[1:2,]
    Q = diag(c(par[2], par[3], 0, 0))
    R = diag(c(par[4], par[5]))
    
    # Some data sets have first (few) out-of-sample timepoints missing, 
    # so get use first non-missing timepoint to initialise
    first_complete_obs = which(!is.na(y_oos[,1]))[1]
    x0_oos = c(y_oos[first_complete_obs,], 
               y_oos[first_complete_obs,])
    P0 = diag(rep(0, 4))

    specify_SSM(
      state_transition_matrix = Phi,
      state_noise_var = Q,
      observation_matrix = A,
      observation_noise_var = R,
      init_state_mean = x0_oos,
      init_state_var = P0)
  }
  
  # Huber and trimmed methods require the last two elements of the initial state mean to be perturbed by a small amount from the first two elements of the initial state mean
  build_oos_fn_huber_trimmed = function(par) {

    phi_coef = par[1]
    Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
    Phi[1,3] = -phi_coef
    Phi[2,4] = -phi_coef
    Phi[3,1] = 1
    Phi[4,2] = 1

    A = diag(4)[1:2,]
    Q = diag(c(par[2], par[3], 0, 0))
    R = diag(c(par[4], par[5]))
    
    # Some data sets have first (few) out-of-sample timepoints missing, 
    # so get use first non-missing timepoint to initialise
    first_complete_obs = which(!is.na(y_oos[,1]))[1]
    second_complete_obs = which(!is.na(y_oos[,1]))[2]
    diff = sum(abs(y_oos[first_complete_obs,] - y_oos[second_complete_obs,]))
    x0_oos = c(y_oos[first_complete_obs,], 
               y_oos[first_complete_obs,] + diff * 0.1)
    P0 = diag(rep(0, 4))

    specify_SSM(
      state_transition_matrix = Phi,
      state_noise_var = Q,
      observation_matrix = A,
      observation_noise_var = R,
      init_state_mean = x0_oos,
      init_state_var = P0)
  }
  
  classical_oos = oos_filter(
    y_oos = y_oos, 
    model = classical_model, 
    build = build_oos_fn)
  
  roams_fut_oos = oos_filter(
    y_oos = y_oos, 
    model = roams_model, 
    build = build_oos_fn,
    multiplier = 2)
  
  oos_outliers_flagged = which(roams_fut_oos$outliers_flagged == 1)
  
  roams_kalman_oos = oos_filter(
    y_oos = y_oos, 
    model = roams_model, 
    build = build_oos_fn,
    multiplier = 1, threshold = Inf)
  
  huber_oos = oos_filter(
    y_oos = y_oos, 
    model = huber_model, 
    build = build_oos_fn_huber_trimmed)
  
  trimmed_oos = oos_filter(
    y_oos = y_oos, 
    model = trimmed_model, 
    build = build_oos_fn_huber_trimmed)
  
  results = results |> 
    add_row(
      name = names(data_sets)[i],
      data_set = list(data_sets[[i]]),
      y_train = list(y_train),
      y_oos = list(y_oos),
      
      classical_model = list(classical_model),
      roams_model = list(roams_model),
      roams_model_list = list(roams_model_list),
      huber_model = list(huber_model),
      trimmed_model = list(trimmed_model),
      
      classical_oos = list(classical_oos),
      roams_fut_oos = list(roams_fut_oos),
      roams_kalman_oos = list(roams_kalman_oos),
      huber_oos = list(huber_oos),
      trimmed_oos = list(trimmed_oos),
      
      oos_outliers_flagged = list(oos_outliers_flagged)
    )
  
  cat("Data set", i, "complete\n")
}

results = results |> 
  mutate(frequency = frequencies)

readr::write_rds(results, "data/results.rds")
```

```{r eval = TRUE}
# load pre-computed results
results = read_rds("data/results.rds")
```

## Results

### Table of MSFEs per animal

This table forms part of Table 3 of @shankar2025. The MSFEs are reported relative to the classical method.

```{r}
results |> 
  select(-roams_model_list) |> 
  mutate(roams_kalman_model = roams_model,
         roams_fut_model = roams_model) |> 
  select(-data_set, -y_train, -frequency, -roams_model) |> 
  rename(y_test = y_oos) |> 
  pivot_longer(cols = ends_with("_oos"), 
               names_to = "method_oos", values_to = "oos_output") |> 
  pivot_longer(cols = ends_with("_model"), 
               names_to = "method_insample", values_to = "insample_output") |> 
  mutate(method_oos = str_remove(method_oos, "_oos"),
         method_insample = str_remove(method_insample, "_model")) |> 
  filter(method_oos == method_insample) |> 
  rename(method = method_oos) |> 
  select(-method_insample) |> 
  mutate(
    MSFE = map2_dbl(
      oos_output, y_test,
      function(oos_output, y_test) {
        SFEs = rowSums((y_test - oos_output$predicted_observations)^2)
        MSFE = mean(SFEs, na.rm = TRUE)
        return(MSFE)
      }
    ),
    MSFE_clean = pmap_dbl(
      list(oos_output, oos_outliers_flagged, y_test),
      function(oos_output, oos_outliers_flagged, y_test) {
        SFEs = rowSums((y_test - oos_output$predicted_observations)^2)
        if (length(oos_outliers_flagged) == 0) {
          MSFE = mean(SFEs, na.rm = TRUE)
        } else {
          MSFE = mean(SFEs[-oos_outliers_flagged], na.rm = TRUE)
        }
        return(MSFE)
      }
    ),
    phi_estimate = map_dbl(insample_output, ~ .$par[1])
  ) |> 
  select(name, method, phi_estimate, MSFE, MSFE_clean) |> 
  group_by(name) |> 
  mutate(
    MSFE = MSFE / MSFE[which(method == "classical")],
    MSFE_clean = MSFE_clean / MSFE_clean[which(method == "classical")]) |>
  knitr::kable(digits = 2)
```

### Table of data set information

This table forms part of Table 2 of @shankar2025. Other ROAMS fit-related information is displayed as well, such as `lambda` chosen and number of `iterations`.

```{r}
results |> 
  select(-roams_model_list) |> 
  mutate(
    n = map_dbl(y_train, ~ nrow(.)),
    n_complete = map_dbl(y_train, ~ sum(!is.na(.[,1]))),
    n_oos = map_dbl(y_oos, ~ nrow(.)),
    n_oos_complete = map_dbl(y_oos, ~ sum(!is.na(.[,1]))),
    
    lambda = map_dbl(roams_model, ~ .$lambda),
    iterations = map_dbl(roams_model, ~ .$iterations),
    
    in_sample_detected = map2_dbl(roams_model, n_complete,
                                  ~ .x$prop_outlying * .y),
    oos_detected = map_dbl(roams_fut_oos,
                           ~ sum(.$outliers_flagged)),
    
    roams_phi = map_dbl(roams_model, ~ .$par[1]),
    classical_phi = map_dbl(classical_model, ~ .$par[1])
    )  |> 
  select(name, frequency, lambda, iterations,
         n, n_complete, in_sample_detected,
         n_oos, n_oos_complete, oos_detected) |> 
  knitr::kable(digits = 3)
```

### Plot outlier diagnostics {.tabset}

#### Cat

```{r}
results |> 
  filter(name == "cat_mia") |> 
  pull(roams_model_list) |> 
  pluck(1) |> 
  autoplot()
```

#### Cow

```{r}
results |> 
  filter(name == "cow_11") |> 
  pull(roams_model_list) |> 
  pluck(1) |> 
  autoplot()
```

#### Polar bear

```{r}
results |> 
  filter(name == "polar_low_freq") |> 
  pull(roams_model_list) |> 
  pluck(1) |> 
  autoplot()
```

#### Seal

```{r}
results |> 
  filter(name == "seal") |> 
  pull(roams_model_list) |> 
  pluck(1) |> 
  autoplot()
```

#### Blue whale

```{r}
results |> 
  filter(name == "whale_2008") |> 
  pull(roams_model_list) |> 
  pluck(1) |> 
  autoplot()
```

### Plot out-of-sample squared forecast errors

Box plots of squared forecast errors for the classical and ROAMS-FUT methods. 

```{r oos-sfe-boxplots}
results |> 
  select(-roams_model_list) |> 
  mutate(
    classical_SFEs = map2(
      classical_oos, y_oos, 
      function(classical_oos, y_oos) {
        SFEs = rowSums((y_oos - classical_oos$predicted_observations)^2)
        SFEs = na.omit(SFEs)
        return(SFEs)
      }
    ),
    roams_SFEs = map2(
      roams_fut_oos, y_oos, 
      function(roams_fut_oos, y_oos) {
        SFEs = rowSums((y_oos - roams_fut_oos$predicted_observations)^2)
        SFEs = na.omit(SFEs)
        return(SFEs)
      }
    )
  ) |> 
  select(name, classical_SFEs, roams_SFEs) |> 
  pivot_longer(c(classical_SFEs, roams_SFEs), 
               names_to = "method", values_to = "SFE") |> 
  mutate(method = ifelse(method == "roams_SFEs",
                         "ROAMS-FUT",
                         "classical")) |> 
  unnest_longer(SFE) |> 
  ggplot() +
  aes(x = method, y = SFE, colour = name) +
  geom_boxplot() +
  facet_wrap(~ name, scales = "free") +
  theme_bw() +
  theme(legend.position = "none")
```

### Plot in-sample fit

```{r in-sample-fits, fig.height = 7}
plots = list()
for (i in 1:length(data_sets)) {
  # Work out x/y ranges with some padding
  x_range <- range(data_sets[[i]]$x, na.rm = TRUE)
  y_range <- range(data_sets[[i]]$y, na.rm = TRUE)
  
  x_pad <- max(diff(y_range) - diff(x_range), 0) / 2  + diff(x_range) * 0.05
  y_pad <- max(diff(x_range) - diff(y_range), 0) / 6  + diff(y_range) * 0.05
  
  result = results |> 
    slice(i)
  
  fitted_paths = tibble(
    x_roams = result$roams_model[[1]]$smoothed_observations[,1],
    y_roams = result$roams_model[[1]]$smoothed_observations[,2],
    x_classical = result$classical_model[[1]]$smoothed_observations[,1],
    y_classical = result$classical_model[[1]]$smoothed_observations[,2]
  )

  detected_outliers = which(rowSums(abs(result$roams_model[[1]]$gamma)) != 0)
  
  plots[[i]] = data_sets[[i]] |> 
    # Retain in-sample data points
    slice(1:round(0.9*n())) |> 
    ggplot() +
    # Country boundaries
    geom_sf(data = world, fill = "grey95", colour = "grey70", size = 0.3) +
    
    # Animal path and points
    geom_point(aes(x = x, y = y), alpha = 0.3) +
    geom_path(data = fitted_paths,
              aes(x = x_classical,
                  y = y_classical), colour = "royalblue") +
    geom_path(data = fitted_paths,
              aes(x = x_roams,
                  y = y_roams), colour = "orange") +
    geom_point(data = data_sets[[i]] |> 
                 slice(1:round(0.9*n())) |> 
                 slice(detected_outliers),
               aes(x = x, y = y), 
               colour = "orange", size = 2, stroke = 1, pch = 1) +
    
    # Focus map on animal’s range
    coord_sf(
      xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
      ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
      expand = FALSE
    ) +
    
    # Labels and theme
    labs(x = "Longitude", y = "Latitude", subtitle = names(data_sets)[i]) +
    theme_minimal() +
    theme(legend.position = "none",
          aspect.ratio = 1)
}

wrap_plots(plots) + plot_layout(ncol = 3)
```

### Plot one-step-ahead out-of-sample predictions

These plots only show the out of sample predictions (not the full path). It is interesting to see where the test set observations lie relative to the training set observations (blank canvas for some of these data sets).

```{r one-step-ahead-preds, fig.height = 7}
plots = list()
for (i in 1:length(data_sets)) {
  # Work out x/y ranges with some padding
  x_range <- range(data_sets[[i]]$x, na.rm = TRUE)
  y_range <- range(data_sets[[i]]$y, na.rm = TRUE)
  
  x_pad <- max(diff(y_range) - diff(x_range), 0) / 2  + diff(x_range) * 0.05
  y_pad <- max(diff(x_range) - diff(y_range), 0) / 6  + diff(y_range) * 0.05
  
  result = results |> 
    slice(i)
  
  oos_paths = tibble(
    x_roams = result$roams_fut_oos[[1]]$predicted_observations[,1],
    y_roams = result$roams_fut_oos[[1]]$predicted_observations[,2],
    x_classical = result$classical_oos[[1]]$predicted_observations[,1],
    y_classical = result$classical_oos[[1]]$predicted_observations[,2]
  )
  
  detected_outliers = which(result$roams_fut_oos[[1]]$outliers_flagged == 1)
  
  plots[[i]] = data_sets[[i]] |> 
    # Retain out-of-sample data points only
    slice((round(0.9*n()) + 1) : n()) |> 
    ggplot() +
    # Country boundaries
    geom_sf(data = world, fill = "grey95", colour = "grey70", size = 0.3) +
    
    # Animal path and points
    geom_point(aes(x = x, y = y), alpha = 0.3) +
    geom_path(data = oos_paths,
              aes(x = x_classical,
                  y = y_classical), colour = "royalblue") +
    geom_path(data = oos_paths,
              aes(x = x_roams,
                  y = y_roams), colour = "orange") +
    geom_point(data = data_sets[[i]] |> 
                 slice((round(0.9*n()) + 1) : n()) |> 
                 slice(detected_outliers),
               aes(x = x, y = y), 
               colour = "orange", size = 2, stroke = 1, pch = 1) +
    
    # Focus map on animal's range
    coord_sf(
      xlim = c(x_range[1] - x_pad, x_range[2] + x_pad),
      ylim = c(y_range[1] - y_pad, y_range[2] + y_pad),
      expand = FALSE
    ) +
    
    # Labels and theme
    labs(x = "Longitude", y = "Latitude", subtitle = names(data_sets)[i]) +
    theme_minimal() +
    theme(legend.position = "none",
          aspect.ratio = 1)
}

wrap_plots(plots) + plot_layout(ncol = 3)
```

## Blue whale plots

This section reproduces the outputs in Section 5.2 of @shankar2025.

### Map 

Fit ROAMS and classical models to blue whale data:

```{r}
y = whale_2008 %>% 
  select(x, y) %>% 
  as.matrix()
```

```{r eval = FALSE}
var_est = c(mad(diff(y[,1]), na.rm = TRUE)^2,
            mad(diff(y[,2]), na.rm = TRUE)^2)
  
build_fn = function(par) {
  
  phi_coef = par[1]
  Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
  Phi[1,3] = -phi_coef
  Phi[2,4] = -phi_coef
  Phi[3,1] = 1
  Phi[4,2] = 1
  
  A = diag(4)[1:2,]
  Q = diag(c(par[2], par[3], 0, 0))
  R = diag(c(par[4], par[5]))
  
  x0 = c(y[1,], y[1,])
  P0 = diag(rep(0, 4))
  
  specify_SSM(
    state_transition_matrix = Phi,
    state_noise_var = Q,
    observation_matrix = A,
    observation_noise_var = R,
    init_state_mean = x0,
    init_state_var = P0)
}

classical_model = classical_SSM(
  y,
  c(0.5, var_est, var_est),
  build_fn,
  lower = c(0, rep(1e-12, 4)),
  upper = c(1, rep(Inf, 4)),
  )

roams_model_list = roams_SSM(
  y,
  c(0.5, var_est, var_est),
  build_fn,
  num_lambdas = 50,
  cores = 4,
  lower = c(0, rep(1e-12, 4)),
  upper = c(1, rep(Inf, 4)),
  B = 50
  )

classical_model = attach_insample_info(classical_model)
roams_model = best_BIC_model(roams_model_list)
roams_model = attach_insample_info(roams_model)

plot_data = tibble(
  x_data = y[,1],
  y_data = y[,2],
  x_roams = roams_model$smoothed_observations[,1],
  y_roams = roams_model$smoothed_observations[,2],
  x_classical = classical_model$smoothed_observations[,1],
  y_classical = classical_model$smoothed_observations[,2],
  outliers = (rowSums(abs(roams_model$gamma)) != 0),
)

readr::write_rds(plot_data, "data/whale_plot_data.rds")
readr::write_rds(roams_model_list, "data/whale_roams_model_list.rds")
readr::write_rds(classical_model, "data/whale_classical_model.rds")
```

Load `whale_plot_data.rds` and `whale_roams_model_list.rds`:

```{r}
plot_data = readr::read_rds("data/whale_plot_data.rds")
roams_model_list = readr::read_rds("data/whale_roams_model_list.rds")
classical_model = readr::read_rds("data/whale_classical_model.rds")
```

```{r}
theme_set(
  theme_bw(base_size = 14) + 
    theme(legend.position = "bottom",
          legend.key.width = unit(1.2, "cm"))
)
```

```{r}
world = ne_countries(scale = "medium", returnclass = "sf")

# States/Provinces
states = ne_states(country = "United States of America")

plot_features = c("observations", "detected outliers", "classical path", "roams path")
colours = c("black", "orange", "royalblue", "orange")
labels = c("Observations", "Detected outliers", "Classical path", "ROAMS path")

p1 = ggplot(plot_data) +
  geom_sf(data = world, fill = "white", color = "grey", linewidth = 0.3) +

  # State/province labels
  geom_sf_text(data = states, aes(label = str_to_upper(name)), size = 3,
               color = "darkgrey", fontface = "bold") +
  geom_sf(data = states, fill = NA, color = "grey", linewidth = 0.3) +

  geom_point(aes(x = x_data, y = y_data, colour = "observations"), 
             size = 2) +
  geom_point(data = plot_data |> filter(outliers), 
             aes(x = x_data, y = y_data, colour = "detected outliers"),
             size = 3, pch = 1, stroke = 1) +
  geom_path(aes(x = x_roams, y = y_roams, 
                colour = "roams path"),
            linewidth = 0.8) +
  geom_path(aes(x = x_classical, y = y_classical, 
                colour = "classical path"), 
            linewidth = 0.8) +
  coord_sf(xlim = c(min(plot_data$x_data, na.rm = TRUE) - 3,
                    max(plot_data$x_data, na.rm = TRUE) + 3),
           ylim = c(min(plot_data$y_data, na.rm = TRUE) - 2,
                    max(plot_data$y_data, na.rm = TRUE) + 2),
           expand = FALSE) +
  theme(panel.grid.major = element_blank(),
        panel.background = element_rect(fill = "#ddecf0")) +
  labs(x = NULL, y = NULL, colour = NULL) +
  scale_colour_manual(values = setNames(colours, plot_features),
                      labels = setNames(labels, plot_features),
                      breaks = plot_features) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))

p1
```

### Map + BIC and proportion outlying curves

Figure 7 of @shankar2025:

```{r fig.width = 9}
p2 = roams_model_list %>% autoplot()

# Extract individual plots for editing
p21 = p2[[1]]
p22 = p2[[2]] +
  labs(y = "Detected outliers") +
  scale_y_continuous(labels = scales::percent)

# Re-combine plots
p2 = p21 + p22 + plot_layout(nrow = 2, axes = "collect")

# Final plot
p1 + p2
```

### Table of parameter estimates

Table 4 of @shankar2025:

```{r}
roams_model = best_BIC_model(roams_model_list)
roams_model = attach_insample_info(roams_model)

tibble(
  roams_model$par,
  classical_model$par,
) |> 
  t() |> 
  as_tibble() |> 
  setNames(c("phi", 
             "state error var lon", "state error var lat",
             "obs error var lon", "obs error var lat"
             )) |> 
  mutate(method = c("ROAMS", "classical")) |> 
  select(method, everything()) |> 
  knitr::kable(digits = 3)
```

### $\|\gamma\|_2$ path plot

```{r}
roams_model_list |> 
  get_attribute("gamma") |> 
  map(~ apply(., 1, function(gamma) {sqrt(sum(gamma^2))})) |> 
  as.data.frame() |> 
  setNames(paste0("lambda_", 1:50)) |>
  tibble() |> 
  mutate(timepoint = 1:122) |> 
  pivot_longer(-timepoint, names_to = "lambda", values_to = "gamma_L2") |> 
  mutate(lambda = as.numeric(str_remove(lambda, "lambda_"))) |> 
  (\(df) 
  ggplot(df) +
  aes(x = lambda, y = gamma_L2, colour = factor(timepoint)) +
  geom_line() +
  geom_text(data = df |> 
              group_by(timepoint) |> 
              slice(1) |> 
              ungroup() |> 
              slice_max(gamma_L2, n = 4),
            aes(label = timepoint)) +
  theme(legend.position = "none"))()
```





#### References


