---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
csl: apa-old-doi-prefix.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

In this vignette, we provide a summary of the main R functions contained in the `roams` package and how to use them. We’ll begin by simulating a data set, fitting a model, inspecting results, and finally using the model to perform out-of-sample (OOS) filtering.


```{r, echo=FALSE, fig.cap="Overview of `roams` functions", fig.align='center', out.width="90%"}
magick::image_read_pdf("data/roams_functions.pdf")
```


We start by loading the required packages and setting a simple ggplot2 theme for consistent plotting.

```{r setup}
library(roams)
library(tidyverse)
library(patchwork)

theme_set(
  theme_bw(base_size = 14) + 
    theme(legend.position = "bottom",
          legend.key.width = unit(1.2, "cm"))
)
```

## Simulating data

`roams` provides a helper function `simulate_data_study1()` which generates example data sets used in the simulations results of Study 1 in @shankar2025. We use this function to simulate a single sample (`samples = 1`) for each of the different sample sizes and outlier settings, and extract specifically the *cluster* setting with $n=100$ timepoints.

```{r}
data = simulate_data_study1(samples = 1, seed = 12345) |> 
  filter(n == 100, setting == "cluster")
```

The data set contains a column `y` which contains an observation in each row:

```{r}
y = data$y[[1]]
head(y)
```

As you can see, the observations are 2-dimensional. You can think of each observation as a longitude/latitude coordinate of an object moving through 2D space. We can plot these observations indexed by time: 

```{r}
plot_data = tibble(
  y1 = y[,1],
  y2 = y[,2],
  timepoint = 1:100
) 

ggplot(plot_data) +
  aes(x = y1, y = y2, colour = timepoint) +
  geom_point() +
  scale_colour_viridis_c()
```

The object first starts near the coordinate $[0,0]$, then it moves in the positive $[y_1, y_2]$ direction, and then it loops around before moving back in the the negative $[y_1, y_2]$ direction. In the top right, there are a few observations scattered away from the main trajectory taken by the object. We suspect that these points are outliers.

## Fitting a ROAMS state-space model

Here are the state and observation equations for a linear Gaussian state-space model:

$$
\begin{align}
\mathbf{y}_t &= \mathbf{A} \mathbf{x}_t + \mathbf{v}_t \\
\mathbf{x}_t &= \mathbf{\Phi} \mathbf{x}_{t-1} + \mathbf{w}_t,
\end{align}
$$

where $\mathbf{y}_t$ is the observation process, $\mathbf{x}_t$ is the state process, $\mathbf{v}_t \sim N(\mathbf{0}, \mathbf{\Sigma}_{\mathbf{v}})$ and $\mathbf{w}_t \sim N(\mathbf{0}, \mathbf{\Sigma}_{\mathbf{w}})$. The state-space model we want to use to model our data is the first-differenced correlated random walk (DCRW) model. The DCRW model has a 2-dimensional $\mathbf{y}_t$ and a 4-dimensional $\mathbf{x}_t$ because $\mathbf{x}_t$ tries to capture the true location of the object at times $t$ and $t-1$ (see @shankar2025 for more details). The matrices in the DCRW model are given by

$$
\begin{align}
\mathbf{\Phi} &= 
\left[
\begin{matrix}
1+\phi&0&-\phi&0 \\
0&1+\phi&0&-\phi \\
1&0&0&0 \\
0&1&0&0 \\
\end{matrix}
\right] \\
\mathbf{A} &= 
\left[
\begin{matrix}
1&0&0&0 \\
0&1&0&0
\end{matrix}
\right],
\end{align}
$$

and the error variance matrices are $\mathbf{\Sigma}_{\mathbf{v}} = \text{diag}(\sigma^2_{\mathbf{v},1}, \sigma^2_{\mathbf{v},2})$ and $\mathbf{\Sigma}_{\mathbf{w}} = \text{diag}(\sigma^2_{\mathbf{w},1}, \sigma^2_{\mathbf{w},2}, 0, 0)$. We assume at time $t=0$, that $\mathbf{x}_0 \sim N(\mathbf{\mu}_0, \mathbf{P}_0)$, where $\mathbf{\mu}_0 = [0,0,0,0]$ and $\mathbf{P}_0 = \text{diag}(10^2, 10^2, 10^2, 10^2)$.

There are a total of 5 unknown parameters in the DCRW model: $\mathbf{\theta} := [\phi,\sigma^2_{\mathbf{w},1}, \sigma^2_{\mathbf{w},2}, \sigma^2_{\mathbf{v},1}, \sigma^2_{\mathbf{v},2}]$. To specify this model, we define a function `build_fn()` that takes in $\mathbf{\theta}$ as an argument and returns $\mathbf{x}_0, \mathbf{P}_0, \mathbf{\Phi}, \mathbf{A}, \mathbf{\Sigma}_{\mathbf{v}}, \mathbf{\Sigma}_{\mathbf{w}}$. The helper function `specify_SSM()` is used inside `build_fn()` to help return these values in the correct format.

```{r}
build_fn = function(par) {
  
  phi_coef = par[1]
  Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
  Phi[1,3] = -phi_coef
  Phi[2,4] = -phi_coef
  Phi[3,1] = 1
  Phi[4,2] = 1
  
  A = diag(4)[1:2,]
  Sigma_w = diag(c(par[2], par[3], 0, 0))
  Sigma_v = diag(c(par[4], par[5]))
  
  mu0 = c(0,0,0,0)
  P0 = diag(rep(10^2, 4))
  
  specify_SSM(
    state_transition_matrix = Phi,
    state_noise_var = Sigma_w,
    observation_matrix = A,
    observation_noise_var = Sigma_v,
    init_state_mean = mu0,
    init_state_var = P0)
}
```

To fit the model using the ROAMS procedure, we use the function `roams_SSM()` and pass in the observations `y` as well as our model specification function `build_fn()`. Since $\phi \in[0,1]$ and $\sigma^2_{\mathbf{w},1}, \sigma^2_{\mathbf{w},2}, \sigma^2_{\mathbf{v},1}, \sigma^2_{\mathbf{v},2} > 0$, we add in these parameter box constraints via the `lower` and `upper` arguments. We provide initial parameter values $[0.5, 1, 1, 1, 1]$ and we set `num_lambdas = 20` to let `roams_SSM()` automatically select a sequence of 20 $\lambda$ values to fit candidate models.

```{r eval = FALSE}
model_list = roams_SSM(
  y = y, 
  init_par = c(0.5, rep(1, 4)), 
  build = build_fn, 
  num_lambdas = 20, 
  lower = c(0, rep(1e-12, 4)), 
  upper = c(1, rep(Inf, 4))
)
```

```{r echo = FALSE}
model_list = readr::read_rds("data/quick_start_model_list.rds")
```

```{r}
model_list
```

## Inspecting the models

Once the models have been fit and assigned to the variable `model_list`, we can plot BIC and proportion-outlying curves across the values of $\lambda$ using the `autoplot()` function:

```{r}
autoplot(model_list)
```

The exact BIC or proportion-outlying values can be obtained using the `get_attribute()` function: 
```{r}
get_attribute(model_list, "prop_outlying")
```

To extract the model corresponding to the lowest BIC, we can use the `best_BIC_model()` function. We assign this model to the variable `final_model`:

```{r}
final_model = best_BIC_model(model_list)
final_model
```

As shown above, printing `final_model` provides a brief summary of the model. It tells us which $\lambda$ value was selected, the proportion of timepoints detected as outlying, the BIC, log-likelihood, and the number of ROAMS iterations.

Other model-specific attributes such as `par` (parameter estimates) can be viewed using `$`:

```{r}
final_model$par
```

By default, more detailed information such as estimates for the hidden states are not included in the model object to save memory. However, we need this information to plot the estimated states. To get this information, we can use the `attach_insample_info()` function:

```{r}
final_model = attach_insample_info(final_model)
```

Below, we use our `final_model` to plot the `smoothed' estimated trajectory of the moving object as well as circle the observations detected as outliers in red.

```{r}
outliers = rowSums(abs(final_model$gamma)) != 0

plot_data = tibble(
  y1 = y[,1],
  y2 = y[,2],
  y1_model = final_model$smoothed_observations[,1],
  y2_model = final_model$smoothed_observations[,2],
  outliers = outliers,
  timepoint = 1:100
)

ggplot(plot_data) +
  aes(x = y1, y = y2, colour = timepoint) +
  geom_point(alpha = 0.3) +
  geom_path(aes(x = y1_model, y = y2_model), 
            linewidth = 1) +
  geom_point(data = plot_data |> filter(outliers),
             colour = "red", pch = 1) +
  scale_colour_viridis_c()
```

The fit looks good and the suspected outliers have been correctly detected.

## Out-of-sample (OOS) filtering

The simulated data set that we generated contains another column named `y_oos`, which contains $n_{\text{oos}} = 20$ out-of-sample observations. We will use our `final_model` to obtain filtered values for `y_oos`.

The out-of-sample data has the same characteristics as the in-sample data, except that it starts where the in-sample data ends. We re-define our `build_fn()` to set the initial state mean $\mathbf{\mu}_0$ to be our best guess of the hidden state at the last in-sample timepoint, $t=100$:

```{r}
final_model$smoothed_states[100,]
```

```{r}
y_oos = data$y_oos[[1]]

build_fn_oos = function(par) {
  
  phi_coef = par[1]
  Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
  Phi[1,3] = -phi_coef
  Phi[2,4] = -phi_coef
  Phi[3,1] = 1
  Phi[4,2] = 1
  
  A = diag(4)[1:2,]
  Sigma_w = diag(c(par[2], par[3], 0, 0))
  Sigma_v = diag(c(par[4], par[5]))
  
  mu0 = final_model$smoothed_states[100,]
  P0 = diag(rep(10^2, 4))
  
  specify_SSM(
    state_transition_matrix = Phi,
    state_noise_var = Sigma_w,
    observation_matrix = A,
    observation_noise_var = Sigma_v,
    init_state_mean = mu0,
    init_state_var = P0)
}
```

Using this new state-space model specification, `build_fn_oos`, we can do out-of-sample filtering using the `oos_filter()` function. We supply data `y_oos`, the model `final_model` (which is needed for its parameter estimates), and our function `build_fn_oos`:

```{r}
result = oos_filter(y_oos, final_model, build_fn_oos)

head(result$filtered_observations)
```

Below, we plot the out-of-sample observations alongside our filtered values. The fit is pretty good:

```{r}
plot_data = tibble(
  y1 = y_oos[,1],
  y2 = y_oos[,2],
  y1_filter = result$filtered_observations[,1],
  y2_filter = result$filtered_observations[,2],
  timepoint = 1:20
)

ggplot(plot_data) +
  aes(x = y1, y = y2, colour = timepoint) +
  geom_point(alpha = 0.3) +
  geom_path(aes(x = y1_filter, y = y2_filter), 
            linewidth = 1) +
  scale_colour_viridis_c()
```

Notice that there are no visible outliers in the out-of-sample data. Let's see what happens to our filtered values if there is an outlier. We can contaminate the 10$^\text{th}$ observation to be outlying:

```{r}
y_oos[10,] = y_oos[10,] + c(10,10)
y_oos[10,]
```

If the model passed to `oos_filter()` is of class `roams_SSM`, then by default the `oos_filter()` function uses the fast-updating threshold (FUT) filter (see @shankar2025) to do out-of-sample filtering. The FUT filter provides robustness against out-of-sample outliers. If the usual Kalman filter is desired, we can set the argument `threshold = Inf`. We try both filters:

```{r}
result_FUT = oos_filter(y_oos, final_model, build_fn_oos)
result_kalman = oos_filter(y_oos, final_model, build_fn_oos, threshold = Inf)
```

Below, we plot the filtered values against the out-of-sample data (where the 10$^\text{th}$ point has been contaminated) for both filtering methods:

```{r}
plot_data = tibble(
  y1 = y_oos[,1],
  y2 = y_oos[,2],
  y1_FUT = result_FUT$filtered_observations[,1],
  y2_FUT = result_FUT$filtered_observations[,2],
  y1_kalman = result_kalman$filtered_observations[,1],
  y2_kalman = result_kalman$filtered_observations[,2],
  timepoint = 1:20
)

p_FUT = ggplot(plot_data) +
  aes(x = y1, y = y2, colour = timepoint) +
  geom_point(alpha = 0.3) +
  geom_path(aes(x = y1_FUT, y = y2_FUT), 
            linewidth = 1) +
  ggtitle("FUT") +
  scale_colour_viridis_c()

p_kalman = ggplot(plot_data) +
  aes(x = y1, y = y2, colour = timepoint) +
  geom_point(alpha = 0.3) +
  geom_path(aes(x = y1_kalman, y = y2_kalman), 
            linewidth = 1) +
  ggtitle("Kalman") +
  scale_colour_viridis_c()

p_FUT + p_kalman + plot_layout(guides = "collect", axes = "collect")
```

As you can see, the FUT filter is able to ignore the outlier whereas the Kalman filter gets pulled towards it.

## Summary

In this vignette, we demonstrated the core functionality of the `roams` package through a complete end-to-end example. We began by simulating bivariate time series data using `simulate_data_study1()`, then specified a state-space model (the DCRW model) through the helper function `specify_SSM()`. Using `roams_SSM()`, we fitted a sequence of models across different values of the tuning parameter $\lambda$ and identified the optimal model via the BIC criterion.

We then explored how to inspect fitted models, extract key quantities (such as parameter estimates and detected outliers), and visualise model diagnostics using `autoplot()` and smoothed states using `attach_insample_info()`.

Finally, we illustrated how to use the fitted model for out-of-sample (OOS) filtering, showing that the fast-updating threshold (FUT) filter can effectively resist the influence of outliers compared to the standard Kalman filter.

### References
