---
title: "Simulation study 1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation study 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
csl: apa-old-doi-prefix.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.height = 6,
  comment = "#>"
)
```

> This page shows how the simulation results in section 4.1 of @shankar2025 are generated.

```{r packages}
library(tidyverse)
library(roams) # version 0.5.1
library(future)

## Number of cores to use
num_cores = 8
study = "study1"
seed = paste0("seed", 2025)
data_path = "data/"
scripts_path = "scripts/"
run_sim = FALSE
```

## Simulate data for Study 1

Simulated data is generated using the `simulate_data_study1()` function. We set it to generate `B = 200` data sets for each combination of sample size (100, 200, 500, 1000) and outlier configuration (clean data, fixed distance, multi-level, cluster). Seed is set to 2025.

```{r settings, message = FALSE, eval = run_sim}
true_par = c(0.8, 0.1, 0.1, 0.4, 0.4)
multi_level_distances = c(3,5,7)

data_study1 = simulate_data_study1(
  sample_sizes = c(100, 200, 500, 1000),
  samples = 200,
  n_oos = 20,
  contamination = 0.1,
  distance = 5,
  sd_cluster = 2,
  mean_cluster = c(20, 20),
  multi_level_distances = multi_level_distances,
  phi_coef = true_par[1],
  sigma2_w_lon = true_par[2],
  sigma2_w_lat = true_par[3],
  sigma2_v_lon = true_par[4],
  sigma2_v_lat = true_par[5],
  initial_state = c(0, 0, 0, 0),
  seed = as.numeric(str_remove(seed, "seed"))
)

readr::write_rds(data_study1, 
                 paste0(data_path, "data_", study, "_", seed, ".rds"),
                 compress = "gz")
```

## Fit models for Study 1  {.tabset}

5 models are fit to each data set: classical, oracle, ROAMS, Huber, and trimmed. Each data frame of models can be large in size and takes days to fit, so they are saved in 5 separate `.rds` files.

### Classical SSM fit

```{r classical_model, message = FALSE, eval = run_sim}
plan(multisession, workers = num_cores)
classical_start = Sys.time()
model_fits_classical_study1 = data_study1 |> 
  mutate(
    classical_model = furrr::future_map(y, function(y) {
      
      build = function(par) {
    
        phi_coef = par[1]
        Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
        Phi[1,3] = -phi_coef
        Phi[2,4] = -phi_coef
        Phi[3,1] = 1
        Phi[4,2] = 1
        
        A = diag(4)[1:2,]
        Q = diag(c(par[2], par[3], 0, 0))
        R = diag(c(par[4], par[5]))
        
        x0 = c(y[1,], y[1,])
        P0 = diag(rep(0, 4))
        
        specify_SSM(
          state_transition_matrix = Phi,
          state_noise_var = Q,
          observation_matrix = A,
          observation_noise_var = R,
          init_state_mean = x0,
          init_state_var = P0)
      }
      
      classical_SSM(
        y = y, 
        init_par = c(0.5, rep(c(mad(diff(y[,1]), na.rm = TRUE)^2, 
                                mad(diff(y[,2]), na.rm = TRUE)^2), 2)),
        build = build, 
        lower = c(0, rep(1e-12, 4)), 
        upper = c(1, rep(Inf, 4))
      )}, 
    .progress = TRUE),
    .keep = "none")
classical_end = Sys.time()
classical_end - classical_start
readr::write_rds(model_fits_classical_study1, 
                 paste0(data_path, "classical_", study, "_", seed, ".rds"), 
                 compress = "gz")
plan(sequential)
rm(model_fits_classical_study1)  ## free up memory
```

### Oracle model

```{r oracle_model, message = FALSE, eval = run_sim}
plan(multisession, workers = num_cores)
oracle_start = Sys.time()
model_fits_oracle_study1 = data_study1 |> 
  mutate(
    oracle_model = furrr::future_map2(y, outlier_locs, function(y, outlier_locs) {
      
      build = function(par) {
    
        phi_coef = par[1]
        Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
        Phi[1,3] = -phi_coef
        Phi[2,4] = -phi_coef
        Phi[3,1] = 1
        Phi[4,2] = 1
        
        A = diag(4)[1:2,]
        Q = diag(c(par[2], par[3], 0, 0))
        R = diag(c(par[4], par[5]))
        
        x0 = c(y[1,], y[1,])
        P0 = diag(rep(0, 4))
        
        specify_SSM(
          state_transition_matrix = Phi,
          state_noise_var = Q,
          observation_matrix = A,
          observation_noise_var = R,
          init_state_mean = x0,
          init_state_var = P0)
      }
      
      oracle_SSM(
        y = y,
        init_par = c(0.5, rep(c(mad(diff(y[,1]), na.rm = TRUE)^2, 
                                mad(diff(y[,2]), na.rm = TRUE)^2), 2)),
        build = build, 
        outlier_locs = outlier_locs,
        lower = c(0, rep(1e-12, 4)), 
        upper = c(1, rep(Inf, 4))
      )}, 
    .progress = TRUE),
    .keep = "none")
oracle_end = Sys.time()
oracle_end - oracle_start
readr::write_rds(model_fits_oracle_study1, 
                 paste0(data_path, "oracle_", study, "_", seed, ".rds"),
                 compress = "gz")
plan(sequential)
rm(model_fits_oracle_study1)  ## free up memory
```

### ROAMS

```{r robust_ipod, message = FALSE, eval = run_sim}
plan(multisession, workers = num_cores)

roams_start = Sys.time()
model_fits_roams_study1 = data_study1 |> 
  mutate(
    model_list = furrr::future_map(y, function(y) {
      
      build = function(par) {
    
        phi_coef = par[1]
        Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
        Phi[1,3] = -phi_coef
        Phi[2,4] = -phi_coef
        Phi[3,1] = 1
        Phi[4,2] = 1
        
        A = diag(4)[1:2,]
        Q = diag(c(par[2], par[3], 0, 0))
        R = diag(c(par[4], par[5]))
        
        x0 = c(y[1,], y[1,])
        P0 = diag(rep(0, 4))
        
        specify_SSM(
          state_transition_matrix = Phi,
          state_noise_var = Q,
          observation_matrix = A,
          observation_noise_var = R,
          init_state_mean = x0,
          init_state_var = P0)
      }
      
      roams_SSM(
        y = y, 
        init_par = c(0.5, rep(c(mad(diff(y[,1]), na.rm = TRUE)^2, 
                                mad(diff(y[,2]), na.rm = TRUE)^2), 2)), 
        build = build, 
        num_lambdas = 20, 
        cores = 1,
        lower = c(0, rep(1e-12, 4)), 
        upper = c(1, rep(Inf, 4)),
        B = 50
      )}, 
    .progress = TRUE),
    best_BIC_model = map(model_list, best_BIC_model),
    .keep = "none")
roams_end = Sys.time()
roams_end - roams_start
readr::write_rds(model_fits_roams_study1, 
                 paste0(data_path, "roams_", study, "_", seed, ".rds"), 
                 compress = "gz")
plan(sequential)
rm(model_fits_roams_study1)  ## free up memory
```

### Huber model

```{r huber_model, message = FALSE, eval = run_sim}
plan(multisession, workers = num_cores)
huber_start = Sys.time()
model_fits_huber_study1 = data_study1 |>
  mutate(
    huber_model = furrr::future_map(y, function(y) {
      
      build_huber_trimmed = function(par) {
    
        phi_coef = par[1]
        Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
        Phi[1,3] = -phi_coef
        Phi[2,4] = -phi_coef
        Phi[3,1] = 1
        Phi[4,2] = 1
      
        A = diag(4)[1:2,]
        Q = diag(c(par[2], par[3], 0, 0))
        R = diag(c(par[4], par[5]))
        
        x0 = c(y[1,], y[1,] + 0.01)
        P0 = diag(rep(0, 4))
        
        specify_SSM(
          state_transition_matrix = Phi,
          state_noise_var = Q,
          observation_matrix = A,
          observation_noise_var = R,
          init_state_mean = x0,
          init_state_var = P0)
      }
      
      huber_robust_SSM(
        y = y,
        init_par = c(0.5, rep(c(mad(diff(y[,1]), na.rm = TRUE)^2, 
                                mad(diff(y[,2]), na.rm = TRUE)^2), 2)),
        build = build_huber_trimmed,
        lower = c(0.001, rep(1e-12, 4)),
        upper = c(0.999, rep(Inf, 4))
      )},
    .progress = TRUE),
    .keep = "none")
huber_end = Sys.time()
huber_end - huber_start
readr::write_rds(model_fits_huber_study1, 
                 paste0(data_path, "huber_", study, "_", seed, ".rds"), 
                 compress = "gz")
plan(sequential)
rm(model_fits_huber_study1)  ## free up memory
```

### Trimmed model

```{r trimmed_model, message = FALSE, eval = run_sim}
plan(multisession, workers = num_cores)
trimmed_start = Sys.time()
model_fits_trimmed_study1 = data_study1 |>
  mutate(
    trimmed_model = furrr::future_map(y, function(y) {
      
      build_huber_trimmed = function(par) {
    
        phi_coef = par[1]
        Phi = diag(c(1+phi_coef, 1+phi_coef, 0, 0))
        Phi[1,3] = -phi_coef
        Phi[2,4] = -phi_coef
        Phi[3,1] = 1
        Phi[4,2] = 1
      
        A = diag(4)[1:2,]
        Q = diag(c(par[2], par[3], 0, 0))
        R = diag(c(par[4], par[5]))
        
        x0 = c(y[1,], y[1,] + 0.01)
        P0 = diag(rep(0, 4))
        
        specify_SSM(
          state_transition_matrix = Phi,
          state_noise_var = Q,
          observation_matrix = A,
          observation_noise_var = R,
          init_state_mean = x0,
          init_state_var = P0)
      }
      
      trimmed_robust_SSM(
        y = y,
        init_par = c(0.5, rep(c(mad(diff(y[,1]), na.rm = TRUE)^2, 
                                mad(diff(y[,2]), na.rm = TRUE)^2), 2)),
        build = build_huber_trimmed,
        alpha = 0.1,
        lower = c(0.001, rep(1e-12, 4)),
        upper = c(0.999, rep(Inf, 4))
      )},
    .progress = TRUE),
    .keep = "none")
trimmed_end = Sys.time()
trimmed_end - trimmed_start
readr::write_rds(model_fits_trimmed_study1,
                 paste0(data_path, "trimmed_", study, "_", seed, ".rds"),
                 compress = "gz")
plan(sequential)
rm(model_fits_trimmed_study1)  ## free up memory
```

## Compute evaluation metrics for fitted models

Using the model fits, in-sample and out-of-sample evaluation metrics (sensitivity, specificity, estimation RMSE, one-step-ahead MSFE) are computed. Functions to compute these metrics are sourced from the `scripts/evaluation_functions.R` script file. This process takes a few minutes to run, so the output is saved to the file `data/eval_metrics_study1.rds`. 

```{r eval = run_sim}
classical_study1 = read_rds(paste0(
  data_path, "classical_", study, "_", seed, ".rds"))
roams_study1 = read_rds(paste0(
  data_path, "roams_", study, "_", seed, ".rds"))
oracle_study1 = read_rds(paste0(
  data_path, "oracle_", study, "_", seed, ".rds"))
huber_study1 = read_rds(paste0(
  data_path, "huber_", study, "_", seed, ".rds"))
trimmed_study1 = read_rds(paste0(
  data_path, "trimmed_", study, "_", seed, ".rds"))

data_study1 = read_rds(paste0(
  data_path, "data_", study, "_", seed, ".rds"))
model_fits_study1 = data_study1 |> 
  bind_cols(classical_study1) |> 
  bind_cols(roams_study1) |> 
  bind_cols(oracle_study1) |> 
  bind_cols(huber_study1) |> 
  bind_cols(trimmed_study1)

model_fits_study1 = model_fits_study1 |> 
  select(-model_list) |>
  rename(roams_fut_model = best_BIC_model) |> 
  mutate(roams_kalman_model = roams_fut_model) 

source(paste0(scripts_path, "evaluation_functions.R"))
eval_start = Sys.time()

eval_metrics_study1 = model_fits_study1 |>
  pivot_longer(ends_with("_model"),
               names_to = "method",
               values_to = "model") |>
  mutate(
    MSFE = pmap_dbl(list(model, method, y_oos), .f = get_MSFE),
    TPR = map2_dbl(model, outlier_locs, .f = get_TPR),
    TNR = map2_dbl(model, outlier_locs, .f = get_TNR),
    TPR_small  = map2_dbl(
      model, outlier_levels,
      .f = ~ get_TPR_multilevel(.x, .y, 
                                level = multi_level_distances[1])),
    TPR_medium = map2_dbl(
      model, outlier_levels,
      .f = ~ get_TPR_multilevel(.x, .y, 
                                level = multi_level_distances[2])),
    TPR_large  = map2_dbl(
      model, outlier_levels,
      .f = ~ get_TPR_multilevel(.x, .y, 
                                level = multi_level_distances[3])),
    par_dist_phi       = map_dbl(
      model, .f = ~ get_par_dist_phi(.x, true_par)),
    par_dist_obs_var   = map_dbl(
      model, .f = ~ get_par_dist_obs_var(.x, true_par)),
    par_dist_state_var = map_dbl(
      model, .f = ~ get_par_dist_state_var(.x, true_par))
    ) |> 
  select(-c(model, 
            y_oos, x_oos, y, y_clean, x, 
            outlier_locs, outlier_levels))  ## reduce size of object by removing data set information

eval_end = Sys.time()
eval_end - eval_start
readr::write_rds(eval_metrics_study1, 
                 paste0(data_path, "eval_metrics_study1.rds"))
```

## Results

We illustrate the results contained in `eval_metrics_study1.rds` using plots and tables.

Load packages for plots and tables:

```{r}
library(patchwork)
library(ggtext)
library(knitr)
library(kableExtra)
library(ggbeeswarm)

theme_set(
  theme_bw(base_size = 14) + 
    theme(legend.position = "bottom",
          legend.key.width = unit(1.2, "cm"))
)

LW = 0.8  # linewidths
PS = 2.0  # point sizes
```

Load `eval_metrics_study1.rds`:

```{r load_eval_metrics}
eval_metrics_study1 = read_rds(paste0(data_path, "eval_metrics_study1.rds"))
```

Check number of replicates per sample size $n$ and configuration:

```{r}
eval_metrics_study1 |> 
  filter(method == "roams_fut_model") |> 
  janitor::tabyl(n, setting) |> 
  knitr::kable()
```

### Mean sensitivity and specificity

Figure 2 of @shankar2025:

```{r fig.height = 5}
settings = c("clean data", "cluster", "fixed distance", "multi-level")
labels_settings = c("Clean data", "Cluster", "Fixed distance", "Multi-level")

p1 = eval_metrics_study1 |> 
  filter(method == "roams_fut_model") |> 
  group_by(n, setting) |> 
  summarise(Sensitivity = mean(TPR),
            Specificity = mean(TNR)) |> 
  pivot_longer(c(Sensitivity, Specificity), 
               names_to = "metric", values_to = "value") |> 
  ggplot() +
  aes(x = factor(n), y = value, 
      colour = setting, group = setting, linetype = setting) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ metric, nrow = 1, scales = "fixed", 
             labeller = labeller(metric = c(
               "Sensitivity" = "Mean Sensitivity", 
               "Specificity" = "Mean Specificity"))) +
  labs(x = "n", y = NULL, colour = NULL, linetype = NULL) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5,1)) +
  scale_colour_manual(
    values = setNames(c("#0072B2","#56B4E9","#009E73","#F5C710"), settings),
    labels = setNames(labels_settings, settings)) +
  scale_linetype_manual(
    values = setNames(c("solid", "dashed", "dotted", "dotdash"), settings),
    labels = setNames(labels_settings, settings)) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))

p2 = eval_metrics_study1 |> 
  filter(setting == "multi-level") |>
  filter(method == "roams_fut_model") |> 
  pivot_longer(starts_with("TPR_"), names_to = "level", values_to = "rate") |>
  mutate(level = case_when(
                           level == "TPR_large" ~ "Large",
                           level == "TPR_medium" ~ "Medium",
                           level == "TPR_small" ~ "Small",
                           )) |> 
  group_by(n, level) |> 
  summarise(TPR = mean(rate)) |> 
  mutate(facet_label = "Mean Sensitivity \n(multi-level outliers)") |> 
  ggplot() +
  aes(x = factor(n), y = TPR, 
      colour = level, group = level, linetype = level) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ facet_label) +
  scale_colour_manual(values = c("#5B126EFF","#D24644FF","#FCA108FF")) +
  labs(x = "n", y = NULL, 
       colour = NULL, linetype = NULL) +
  scale_y_continuous(labels = scales::percent, limits = c(0.5,1)) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))

p1 + p2 + plot_layout(ncol = 2, widths = c(2,1), axes = "collect")
```

### Raw sensitivity and specificity {.tabset}

Sensitivity and specificity of the ROAMS method for each replicate are plotted. Each point corresponds to a replicate: 

#### Sensitivity

```{r fig.height = 5}
eval_metrics_study1 |> 
  filter(method == "roams_fut_model") |> 
  filter(setting != "clean data") |> 
  ggplot() +
  aes(x = factor(n), y = TPR) +
  geom_beeswarm(corral = "random", corral.width = 0.9, alpha = 0.5,
                cex = 0.5) +
  facet_wrap(~ setting, nrow = 1, scales = "fixed", 
             labeller = as_labeller(setNames(labels_settings, settings))) +
  labs(x = "n", y = "True positive rate") +
  scale_y_continuous(labels = scales::percent, limits = c(0,1))
```

#### Specificity

```{r fig.height = 5}
eval_metrics_study1 |> 
  filter(method == "roams_fut_model") |> 
  ggplot() +
  aes(x = factor(n), y = TNR) +
  geom_beeswarm(corral = "random", corral.width = 0.9, alpha = 0.5,
                cex = 0.5) +
  facet_wrap(~ setting, nrow = 2, scales = "fixed",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  labs(x = "n", y = "True negative rate") +
  scale_y_continuous(labels = scales::percent, limits = c(0,1))
```

#### Multi-level sensitivity

```{r fig.height = 5}
eval_metrics_study1 |> 
  filter(method == "roams_fut_model") |> 
  filter(setting == "multi-level") |>
  pivot_longer(starts_with("TPR_"), names_to = "level", values_to = "rate") |>
  mutate(level = case_when(
                           level == "TPR_large" ~ "Large",
                           level == "TPR_medium" ~ "Medium",
                           level == "TPR_small" ~ "Small",
                           )) |> 
  ggplot() +
  aes(x = factor(n), y = rate, colour = level) +
  geom_beeswarm(corral = "random", corral.width = 0.9, alpha = 0.5,
                cex = 0.5) +
  facet_wrap(~ level) +
  scale_colour_manual(values = c("#5B126EFF","#D24644FF","#FCA108FF")) +
  labs(x = "n", y = "True positive rate \n (multi-level outliers)") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0,1))
```

### Out-of-sample MSFE {.tabset}

Mean-aggregated (Figure 3 of @shankar2025) and raw box plots of the one-step-ahead out-of-sample MSFE:

#### Mean MSFE

```{r fig.height=6}
methods = c("roams_fut_model", "roams_kalman_model", "classical_model", "huber_model", "trimmed_model", "oracle_model")
labels = c("**ROAMS-FUT**", "**ROAMS-Kalman**", "Classical", "Huber", "Trimmed", "Oracle")
colours = c("orange", "gold", "royalblue", "brown3", "purple3", "green4")
linetypes = c("solid", "solid", "dotted", "dashed", "dashed", "dotdash")

eval_metrics_study1 |> 
  select(sample, n, setting, method, MSFE) |> 
  group_by(n, setting, method) |> 
  summarise(MSFE = mean(MSFE)) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = MSFE, colour = method, linetype = method,
      group = method) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  scale_linetype_manual(values = setNames(linetypes, methods),
                        labels = setNames(labels, methods)) +
  labs(x = "n", y = "Mean MSFE", colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

#### Raw MSFE

```{r}
eval_metrics_study1 |> 
  select(sample, n, setting, method, MSFE) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = MSFE, colour = method) +
  geom_boxplot() +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
    labels = setNames(labels, methods)) +
  labs(x = "n", y = "MSFE", colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

### Phi: Parameter estimation error (RMSE) {.tabset}

Mean-aggregated and raw RMSE of each method's phi estimate from the true value of 0.8:

#### Mean RMSE

```{r}
methods = c("roams_fut_model", "classical_model", "huber_model", "trimmed_model", "oracle_model")
labels = c("**ROAMS**", "Classical", "Huber", "Trimmed", "Oracle")
colours = c("orange", "royalblue", "brown3", "purple3", "green4")
linetypes = c("solid", "dotted", "dashed", "dashed", "dotdash")

eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |> 
  group_by(n, setting, method) |> 
  summarise(RMSE = mean(sqrt(par_dist_phi))) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = RMSE, colour = method, linetype = method,
      group = method) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ setting, scales = "free_y", 
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  scale_linetype_manual(values = setNames(linetypes, methods),
                        labels = setNames(labels, methods)) +
  labs(x = "n", y = "Mean RMSE (phi)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

#### Raw RMSE

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |>
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = sqrt(par_dist_phi), colour = method) +
  geom_boxplot() +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
    labels = setNames(labels, methods)) +
  labs(x = "n", y = "RMSE (phi)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

### State error variances: Parameter estimation error (RMSE) {.tabset}

Mean-aggregated and raw RMSE of each method's state error variances estimates from the true value of (0.1, 0.1):

#### Mean RMSE

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |> 
  group_by(n, setting, method) |> 
  summarise(RMSE = mean(sqrt(par_dist_state_var))) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = RMSE, colour = method, linetype = method,
      group = method) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  scale_linetype_manual(values = setNames(linetypes, methods),
                        labels = setNames(labels, methods)) +
  labs(x = "n", y = "Mean RMSE (state error vars)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

#### Raw RMSE

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = sqrt(par_dist_state_var), colour = method) +
  geom_boxplot() +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  labs(x = "n", y = "RMSE (state error vars)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

### Observation error variances: Parameter estimation error (RMSE) {.tabset}

Mean-aggregated and raw RMSE of each method's observation error variances estimates from the true value of (0.4, 0.4):

#### Mean RMSE

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |> 
  group_by(n, setting, method) |> 
  summarise(RMSE = mean(sqrt(par_dist_obs_var))) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = RMSE, colour = method, linetype = method,
      group = method) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
    labels = setNames(labels, methods)) +
  scale_linetype_manual(values = setNames(linetypes, methods),
  labels = setNames(labels, methods)) +
  labs(x = "n", y = "Mean RMSE (obs error vars)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

#### Raw RMSE

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |>
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = factor(n), y = sqrt(par_dist_obs_var), colour = method) +
  geom_boxplot() +
  facet_wrap(~ setting, scales = "free_y",
             labeller = as_labeller(setNames(labels_settings, settings))) +
  scale_colour_manual(values = setNames(colours, methods),
    labels = setNames(labels, methods)) +
  labs(x = "n", y = "RMSE (obs error vars)", 
       colour = NULL, linetype = NULL) +
  theme(legend.text = element_markdown())
```

### Table of parameter estimation mean RMSEs for n = 1000

Table 2 of @shankar2025. For each outlier configuration, the mean parameter estimation RMSEs of each method is reported relative to the oracle benchmark method:

```{r}
eval_metrics_study1 |> 
  filter(method != "roams_kalman_model") |> 
  filter(n == 1000) |>  
  select(setting, method, par_dist_phi:par_dist_state_var) |> 
  group_by(setting, method) |> 
  summarise(phi = mean(sqrt(par_dist_phi)),
            obs_var = mean(sqrt(par_dist_obs_var)),
            state_var = mean(sqrt(par_dist_state_var))) |> 
  mutate(
    phi = phi / phi[which(method == "oracle_model")],
    obs_var = obs_var / obs_var[which(method == "oracle_model")],
    state_var = state_var / state_var[which(method == "oracle_model")]
    ) |> 
  ungroup() |> 
  pivot_longer(c(phi, obs_var, state_var), names_to = "parameter") |> 
  mutate(parameter = fct_inorder(parameter)) |> 
  mutate(method = str_remove(method, "_model"),
         method = str_to_title(method),
         method = ifelse(method == "Roams_fut", "ROAMS", method)) |> 
  pivot_wider(names_from = method, values_from = value) |> 
  arrange(parameter, setting) |> 
  select(parameter,setting, everything()) |>
  mutate(parameter = case_when(
    parameter == "phi" ~ "phi autocorrelation",
    parameter == "obs_var" ~ "Observation error var",
    parameter == "state_var" ~ "State error var"
    )) |> 
  mutate(setting = str_to_sentence(setting)) |> 
  rename(Parameter = parameter,
         Configuration = setting) |> 
  select(Parameter, Configuration, 
         Oracle, ROAMS, Huber, Trimmed, Classical) |> 
  kable(digits = 2)
```

### References
