---
title: "Simulation study 3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation study 3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
csl: apa-old-doi-prefix.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

> This page shows how the simulation results in Section 4.3 of @shankar2025 are generated.

```{r packages}
library(tidyverse)
library(roams) # version 0.6.5
library(future)

## Number of cores to use
study = "study2"  # Study 3 re-uses data and model fits from Study 2
seed = paste0("seed", 2025.3)
data_path = "data/"
scripts_path = "scripts/"
run_sim = FALSE
```

## Simulate data for Study 3

The code below takes the out-of-sample data from Study 2 and contaminates it with outliers to form the data for Study 3. The models fitted in-sample from Study 2 are also loaded so that performance on contaminated out-of-sample data can be evaluated for Study 3.

```{r eval = run_sim}
classical_study2 = read_rds(paste0(
  data_path, "classical_", study, "_", seed, ".rds"))
roams_study2 = read_rds(paste0(
  data_path, "roams_", study, "_", seed, ".rds"))
oracle_study2 = read_rds(paste0(
  data_path, "oracle_", study, "_", seed, ".rds"))
huber_study2 = read_rds(paste0(
  data_path, "huber_", study, "_", seed, ".rds"))
trimmed_study2 = read_rds(paste0(
  data_path, "trimmed_", study, "_", seed, ".rds"))

data_study2 = read_rds(paste0(
  data_path, "data_", study, "_", seed, ".rds"))
model_fits_study2 = data_study2 |> 
  bind_cols(classical_study2) |> 
  bind_cols(roams_study2) |> 
  bind_cols(oracle_study2) |> 
  bind_cols(huber_study2) |> 
  bind_cols(trimmed_study2)

# Filter to outlier distance = 5 to create model_fits_study3
model_fits_study3 = model_fits_study2 |> 
  select(-model_list) |>
  rename(roams_fut_model = best_BIC_model) |> 
  mutate(roams_kalman_model = roams_fut_model) |> 
  filter(distance == 5) |> 
  select(-y, -x, -y_clean, -outlier_locs, -distance, -x_oos)

oos_outlier_locs = tibble(
  contamination_oos = c(0, 0.05, 0.1, 0.15, 0.2),
  locs = list(NULL, 10, c(5,15), c(5,10,15), c(5,10,15,20))
)

# Generate random vectors for out-of-sample contamination
set.seed(as.numeric(str_remove(seed, "seed")))
samples = 300
angles = runif(samples*4, max = 2*pi)
angles = split(angles, rep(1:samples, each = 4)) |> unname()


oos_outlier_vecs = tibble(
  sample = 1:samples,
  vec = map(angles, function(angle) {map(angle, ~ c(cos(.), sin(.)) * 5)})  # outlier distance = 5
)

# Contaminate out-of-sample data
model_fits_study3 = model_fits_study3 |> 
  mutate(
    y_oos_contaminated = pmap(
      list(sample, contamination, y_oos),
      function(sample_no, contamination, y_oos) {
        locs = oos_outlier_locs |> 
          # Strange issue that simply filtering for contamination_oos == contamination does not work, so we use the following workaround:
          filter(contamination_oos < contamination + 0.001,
                 contamination_oos > contamination - 0.001) |> 
          pull(locs) |> 
          pluck(1)
        
        y_oos_contaminated = y_oos
        for (timepoint in 1:20) {
          if (timepoint %in% locs) {
            
            vec = oos_outlier_vecs |> 
              filter(sample == sample_no) |> 
              pull(vec) |> 
              pluck(1) |> 
              pluck(timepoint / 5)  # get appropriate random vector
            
            y_oos_contaminated[timepoint,] = y_oos_contaminated[timepoint,] + vec
          }
        }
        
        return(y_oos_contaminated)
      }
    )
  )

model_fits_study3 = model_fits_study3 |> 
  mutate(contamination = as.character(contamination)) |> 
  left_join(oos_outlier_locs |> 
              rename(contamination = contamination_oos) |> 
              mutate(contamination = as.character(contamination)), 
            by = "contamination") |> 
  rename(true_oos_outlier_locs = locs) |> 
  mutate(contamination = as.numeric(contamination))
```

## Compute evaluation metrics for fitted models

Using the model fits from Study 2, one-step-ahead contaminated out-of-sample MSFEs are computed against the clean observations `y_oos`. The function to compute this metric, `get_MSFE_contaminated()`, is sourced from the `scripts/evaluation_functions.R` script file. This process takes a few minutes to run, so the output is saved to the file `data/eval_metrics_study3.rds`. 

```{r eval = run_sim}
source(paste0(scripts_path, "evaluation_functions.R"))
eval_start = Sys.time()

eval_metrics_study3 = model_fits_study3 |>
  pivot_longer(ends_with("_model"),
               names_to = "method",
               values_to = "model") |>
  mutate(
    MSFE_contaminated = pmap_dbl(list(
      model, 
      method, 
      y_oos_contaminated, 
      y_oos, 
      true_oos_outlier_locs
      ), .f = get_MSFE_contaminated)
    ) |> 
  select(-c(model, y_oos, y_oos_contaminated))  ## reduce size of object by removing data set information

eval_end = Sys.time()
eval_end - eval_start
readr::write_rds(eval_metrics_study3, 
                 paste0(data_path, "eval_metrics_study3.rds"))
```

## Results

We illustrate the results contained in `eval_metrics_study3.rds` using plots and tables.

Load packages for plots and tables:

```{r}
library(patchwork)
library(ggtext)
library(knitr)
library(kableExtra)
library(ggbeeswarm)

theme_set(
  theme_bw(base_size = 14) + 
    theme(legend.position = "bottom",
          legend.key.width = unit(1.2, "cm"))
)

LW = 0.8  # linewidths
PS = 2.0  # point sizes
```

Load `eval_metrics_study1.rds`:

```{r load_eval_metrics}
eval_metrics_study3 = read_rds(paste0(data_path, "eval_metrics_study3.rds"))
```

Check number of replicates per contamination rate:

```{r}
eval_metrics_study3 |> 
  filter(method == "roams_fut_model") |> 
  janitor::tabyl(contamination) |> 
  knitr::kable()
```

### Contaminated out-of-sample MSFE {.tabset}

Mean-aggregated (Figure 6 of @shankar2025) and raw box plots of the one-step-ahead contaminated out-of-sample MSFE:

#### Mean MSFE

```{r}
methods = c("roams_fut_model", "roams_kalman_model", "classical_model", "huber_model", "trimmed_model", "oracle_model")
labels = c("**ROAMS-FUT**", "**ROAMS-Kalman**", "Classical", "Huber", "Trimmed", "Oracle")
colours = c("orange", "gold", "royalblue", "brown3", "purple3", "green4")
linetypes = c("solid", "solid", "dotted", "dashed", "dashed", "dotdash")

eval_metrics_study3 |> 
  group_by(contamination, method) |> 
  summarise(MSFE_contaminated = mean(MSFE_contaminated)) |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = contamination, y = MSFE_contaminated, 
      colour = method, linetype = method) +
  geom_line(linewidth = LW) +
  geom_point(size = PS) +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  scale_linetype_manual(values = setNames(linetypes, methods),
                        labels = setNames(labels, methods)) +
  labs(x = "Contamination rate (in-sample and out-of-sample)",
       y = "Mean MSFE", colour = NULL, linetype = NULL) +
  scale_x_continuous(labels = scales::percent) +
  theme(legend.text = element_markdown())
```

#### Raw MSFE

The cases where ROAMS-FUT doesn't do well are when it has detected an out-of-sample timepoint as an outlier *just before* a true outlier. This causes the FUT filter to not be able to detect the true outlier, resulting in poor MSFE.

```{r}
eval_metrics_study3 |> 
  mutate(method = factor(method, levels = methods)) |> 
  ggplot() +
  aes(x = contamination, y = MSFE_contaminated, 
      colour = method, group = interaction(contamination, method)) +
  geom_boxplot() +
  scale_colour_manual(values = setNames(colours, methods),
                      labels = setNames(labels, methods)) +
  labs(x = "Contamination rate (in-sample and out-of-sample)",
       y = "MSFE", colour = NULL) +
  scale_x_continuous(labels = scales::percent) +
  theme(legend.text = element_markdown())
```

### References

